---
title: "**DESIGN AND ANALYSIS OF EXPERIMENTS**"
author: "**B.M Njuguna**"
date: "**2022-10-29**"
output:
  pdf_document:
    pandoc_args: "--listings"
    keep_tex: yes
    latex_engine: xelatex
    fig_caption: yes
    highlight: haddock
number_sections: no
toc: no
toc_depth: 4
df_print: paged
geometry: left=2cm,right=2cm,top=2cm,bottom=2cm
header-includes:
- \usepackage{graphicx}
- \usepackage{float}
- \usepackage{amsmath}
- \nocite{*}
- \usepackage[numbers]{natbib}
- \usepackage{titlesec}
- \titleformat{\section}[block]{\color{black}\Large\bfseries\filcenter}{}{1em}{}
- \usepackage{mdframed}
---

```{r setup, include=FALSE}
options(scipen = 1, digits = 3)
library(knitr)
knitr::opts_chunk$set(warning = FALSE, message = FALSE, echo = FALSE, comment = NA)
def_hook <- knit_hooks$get("output")
knit_hooks$set(output = function(x, options){
  out <- def_hook(x, options)
  return(paste("\\begin{mdframed}[backgroundcolor=gray!10]\\begin{verbatim}", x,
               "\\end{verbatim}\\end{mdframed}", collapse = "\n"))
})
library(readxl)
library(knitr)
```

```{=tex}
\newpage
\tableofcontents
\newpage
```
# Introduction

This paper is based on advanced experimental design for scientific Studies. The various formulas used in different designs and experiments are as outlined below.

# 1. One Way Classification Model

This is basically the Analysis of Covariance(ANCOVA). Here single factor experiment with single covariate is considered. The model is as follows;

$$Y_{ij} = \mu+\tau_i+\beta(x_{ij}-\bar{x..} )+ \epsilon_{ij}$$

where; i = 1, 2, ..., t and j = 1,2, ... r

$Y_{ij}$- is the jth response variable taken under the ith treatment

$\mu$ - overall mean

$x_{ij}$ - The measure of covariate corresponding to $Y_{ij}$

$\bar{x..}$ - The mean of $x_{ij}$ value

$\beta$ - The linear regression coefficient of $Y_{ij}$ and $x{ij}$

The notations used are;

$$S_{yy}=\sum_{i=1}^t\sum_{j=1}^r(Y_{ij}-\bar{Y..})^2=\sum_{i=1}^t\sum_{j=1}^rY_{ij}^2-\frac{Y..^2}{tr}$$

$$S_{xx}=\sum_{i=1}^t\sum_{j=1}^r(X_{ij}-\bar{x..})^2=\sum_{i=1}^t\sum_{j=1}^rX_{ij}^2-\frac{X..^2}{tr}$$

$$S_{xy}=\sum_{i=1}^t\sum_{j=1}^r(Y_{ij}-\bar{Y..})(X_{ij}-\bar{x..})=\sum_{i=1}^t\sum_{j=1}^rY_{ij}X_{ij}-\frac{Y..X..}{tr}$$

$$T_{yy}=\sum_{i=1}^t\sum_{j=1}^r(Y_{i.}-\bar{Y..})^2=\frac{\sum_{i=1}^tY_{i.}^2}{r}-\frac{Y_{..}^2}{tr}$$

$$T_{xx}=\sum_{i=1}^t\sum_{j=1}^r(X_{i.}-\bar{X..})^2=\frac{\sum_{i=1}^tX_{i.}^2}{r}-\frac{X_{..}^2}{tr}$$

$$\frac{\sum_{i=1}^tY_{i}Xi}{r}-\frac{Y_{..}X..}{tr}$$

$$E_{yy}=S_{yy}-T_{yy}$$

$$E_{xx}=S_{xx}-T_{xx}$$

$$E_{xy}=S_{xy}-T_{xy}$$

The statistical analysis is;

LSE of $\mu$ is; $\hat{\mu}=\bar{Y..}$

Then;

$$\hat{\beta}=\frac{E_{xy}}{E_{xx}}$$

$$SSE = E_{yy} - \frac{E_{xy}^2}{Exx}$$

SSE usually have t(r-1)-1 degrees of freedom

Now suppose we use to test $\tau_i=0$. Then under Null hypothesis, the reduced model will be;

$$Y_{ij} = \mu+\beta(x_{ij}-\bar{x..} + \epsilon_{y}$$

Then;

$$\hat{\beta}=\frac{S_{xy}}{S_{xx}}$$

And;

$$SSE' = S_{yy}-\frac{S_{xy}^2}{S_{xx}}$$

note $SSE$ is smaller than $SSE'$, where $SSE'-SSE$ is a reduction in sums of squares due to $\tau_i$. Therefore for testing $\tau_i=0$, the test statistic is;

$$F_{calc}=\frac{(SSE'-SSE)/(t-1)}{SSE/(t(r-1)-1)}$$

We test it against $F_{(t-1),t(r-1)-1,\alpha}$. The ANOVA table is as follows;

```{r}
ANOVA <-read_excel("D:/Documents/R-Studio Programms/Design/CovAnova.xlsx")
kableExtra::kable(ANOVA)
```

\newpage

# 2. Nested Designs

## i. Two - Stage Nested Designs

The statistical model is;

$$Y_{ijk}=\mu+\tau_i+\beta_{j(i)}+\epsilon_{(ij)k}$$

where, i = 1,2,...,a, j = 1,2,...,b, and k = i,2,...r

$\mu$ - overall mean

$\tau_i$ - Effect of the ith factor A

$\beta_{j(i)}$ - Effect of the jth factor B nested under factor A

$\epsilon_{(ij)k}$ - Random error term.

The sums of squares are partitioned as follows;

$$SS_{total} = SS_{A}+ss_{B(A)}+SS_{Error}$$

They are calculated as follows;

$$SS_{Total} = \sum_{i=1}^a\sum_{j=1}^b\sum_{k=1}^rY_{ijk}^2-\frac{(Y...)^2}{abr}$$

$$SS_A=\frac{\sum_{i=1}^aY_{i..}^2}{br}-\frac{(Y...)^2}{abr}$$

$$SS_{B(A)} = \frac{\sum_{i=1}^a\sum_{j=1}^bY_{ij.}^2}{r}-\frac{\sum_{i=1}^aYi..^2}{br}$$

$$SS_{Error}=\sum_{i=1}^a\sum_{j=1}^b\sum_{k=1}^rY_{ijk}^2-\frac{\sum_{i=1}^a\sum_{j=1}^bY_{ij.}^2}{r}$$

The ANOVA table is as follows;

+---------------------+----------------------------------+--------------------------------------------------+---------------------------------------------------------------+
| Source of Variation | df                               | SS                                               | MS                                                            |
+:===================:+:================================:+:================================================:+:=============================================================:+
| $$                  | $$                               | $$                                               | $$                                                            |
|            A        |                             a-1  |                                         SS_{A}   |                                                       MS_A    |
|           $$        |                             $$   |                                          $$      |                                                       $$      |
+---------------------+----------------------------------+--------------------------------------------------+---------------------------------------------------------------+
| $$                  | $$a(b-1)$$                       | $$                                               | $$                                                            |
|          B(A)       |                                  |                                       SS_{B(A)}  |                                                    MS_{B(A)}  |
|           $$        |                                  |                                          $$      |                                                       $$      |
+---------------------+----------------------------------+--------------------------------------------------+---------------------------------------------------------------+
| $$                  | $$ab(r-1)$$                      | $$                                               | $$                                                            |
|          Error      |                                  |                                       SS_{Error} |                                                    MS_{Error} |
|           $$        |                                  |                                          $$      |                                                       $$      |
+---------------------+----------------------------------+--------------------------------------------------+---------------------------------------------------------------+
| $$                  | $$                               | $$                                               |                                                               |
|          Total      |                            abr-1 |                                       SS_{Total} |                                                               |
|           $$        |                             $$   |                                          $$      |                                                               |
+---------------------+----------------------------------+--------------------------------------------------+---------------------------------------------------------------+

The appropriate statistic for testing the effect of factor A and B depends on whether the levels of A and B are fixed or random. The Expected Mean Squares in the two stage nested design are as follows;

+-------------+-------------------------------------------------------+---------------------------------------------------+------------------------------------------+
| $$          | A - fixed                                             | A - fixed                                         | A - Random                               |
| E(MS)       |                                                       |                                                   |                                          |
| $$          | B- fixed                                              | B - Random                                        | B - Random                               |
+=============+:=====================================================:+:=================================================:+:========================================:+
| $$          | $$                                                    | $$                                                | $$                                       |
| E(MS_A)     | \frac{\sigma^2+br\sum\tau_i^2}{a-1}                   | \sigma^2+r\sigma^2_\beta+\frac{br\sum\tau_i}{a-1} | \sigma^2+r\sigma^2_\beta+br\sigma^2_\tau |
| $$          | $$                                                    | $$                                                | $$                                       |
+-------------+-------------------------------------------------------+---------------------------------------------------+------------------------------------------+
| $$          | $$                                                    | $$                                                | $$                                       |
| E(MS_{AB})  | \sigma^2+\frac{r\sum_{i}\sum_j\beta_{j(i)}^2}{a(b-1)} | \sigma^2+r\sigma^2_\beta                          | \sigma^2+r\sigma^2_\beta                 |
| $$          | $$                                                    | $$                                                | $$                                       |
+-------------+-------------------------------------------------------+---------------------------------------------------+------------------------------------------+
| $$          | $$                                                    | $$                                                | $$                                       |
| E(MS_E)     | \sigma^2                                              | \sigma^2                                          | \sigma^2                                 |
| $$          | $$                                                    | $$                                                | $$                                       |
+-------------+-------------------------------------------------------+---------------------------------------------------+------------------------------------------+

The testing of hypothesis is as follows;

The testing of hypothesis is as follows;

1.  [**when A is fixed and B is random**]{.underline}

To test $H_0:\tau_i=0$ vs $H_1:\tau_i\ne0$ the test statistic is;

$$F_{calc}=\frac{MS_A}{MS_{B(A)}}$$

We reject $H_0$ is $F_{calc}>F_{a-1,a(b-1),\alpha}$

To test $H_0:\sigma^2_\beta=0$ vs $H_1:\sigma^2_\beta\neq0$, the test statistic is;

$$F{(calc)=\frac{MS_{B(A)}}{MS_{Error}}}$$

We reject $H_0$ is $F_{calc}>F_{a(b-1),ab(r-1),\alpha}$

2.  [**A and B fixed**]{.underline}

To test $H_0:\tau_i=0$ vs $H_1:\tau_i\ne0$ the test statistic is;

$$F_{calc}=\frac{MS_A}{MS_{Error}}$$

To test $H_0:\beta_{j(i)}=0$ vs $H_1:\beta_{j(i)}\neq0$, the test statistic is;

$$F{(calc)=\frac{MS_{B(A)}}{MS_{Error}}}$$

3.  [**A and B random**]{.underline}

To test $H_0:\sigma^2_{\tau}=0$ vs $H_1:\sigma^2{_\tau}\ne0$ the test statistic is;

$$F_{calc}=\frac{MS_A}{MS_{B(A)}}$$

We reject $H_0$ is $F_{calc}>F_{a-1,a(b-1),\alpha}$

To test $H_0:\sigma^2_\beta=0$ vs $H_1:\sigma^2_\beta\neq0$, the test statistic is;

$$F{(calc)=\frac{MS_{B(A)}}{MS_{Error}}}$$

We reject $H_0$ is $F_{calc}>F_{a(b-1),ab(r-1),\alpha}$

## ii. Three Stage Nested Design

The model is written as;

$$Y_{ijkl}=\mu+\tau_i+\beta_{j(i)}+\gamma_{k(ij)}+\epsilon_{(ijk)l}$$

where; i=1,2,...a, j = 1,2,...,b, k = 1,2,...c and l = 1,2,...r

$\mu$ - overall mean

$\tau_i$ - Effect of the ith factor A

$\beta_{j(i)}$ - Effect of the jth factor B nested under factor A

$\gamma_{k(ij)}$ - Effect of the kth factor B nested under factor A and B

$\epsilon_{(ij)k}$ - Random error term.

The sums of squares are partitioned as follows;

$$SS_{Total}=SS_A+SS_{B(A)}+SS_{C(B)}+SS_{Error}$$

$$SS_{Total = \sum_i\sum_j\sum_k\sum_lY_{ijkl}^2}-\frac{Y_{...}^2}{abcr}$$

$$SS_A = \frac{\sum_iY_{i...}^2}{bcr}-CT$$

$$SS_{B(A)}=\frac{\sum_i\sum_jY_{ij..}^2}{cr}-\frac{\sum_iY_{i...}^2}{bcr}$$

$$SS_{C(B)}=\frac{\sum_i\sum_j\sum_kY_{ijk.}^2}{r}-\frac{\sum_i\sum_jY_{ij..}^2}{cr}$$

$$SS_{Error}=\sum_i\sum_j\sum_k\sum_lY_{ijkl}^2-\frac{\sum_i\sum_j\sum_kY_{ijk.}^2}{r}$$

The ANOVA Table is as follows;

+---------------------+----------------------------------+-----------------------------------------------+------------------------------------------------------------+
| Source of Variation | df                               | SS                                            | MS                                                         |
+:===================:+:================================:+:=============================================:+:==========================================================:+
| $$                  | $$                               | $$                                            | $$                                                         |
|            A        |                           a-1    |                                       SS_A    |                                                    MS_A    |
|           $$        |                           $$     |                                       $$      |                                                    $$      |
+---------------------+----------------------------------+-----------------------------------------------+------------------------------------------------------------+
| $$                  | $$                               | $$                                            | $$                                                         |
|          B(A)       |                           b-1    |                                    SS_{B(A)}  |                                                 MS_{B(A)}  |
|           $$        |                           $$     |                                       $$      |                                                    $$      |
+---------------------+----------------------------------+-----------------------------------------------+------------------------------------------------------------+
| $$                  | $$                               | $$                                            | $$                                                         |
|          C(A)       |                         ab(c-1)  |                                    SS_{C(B)}  |                                                 MS_{C(B)}  |
|           $$        |                           $$     |                                       $$      |                                                    $$      |
+---------------------+----------------------------------+-----------------------------------------------+------------------------------------------------------------+
| $$                  | $$                               | $$                                            | $$                                                         |
|          Error      |                         abc(r-1) |                                    SS_{Error} |                                                 MS_{Error} |
|           $$        |                           $$     |                                       $$      |                                                    $$      |
+---------------------+----------------------------------+-----------------------------------------------+------------------------------------------------------------+
| $$                  | $$                               | $$                                            |                                                            |
|          Total      |                          abcr-1  |                                    SS_{Total} |                                                            |
|           $$        |                           $$     |                                       $$      |                                                            |
+---------------------+----------------------------------+-----------------------------------------------+------------------------------------------------------------+

\newpage

# 3. Balanced Incomplete Block Design

In some experiments, the number of treatments is large, the adoption of RCBD therefore may result in an increase of error variance due to large block size. Here we use the BIBD. The parameters of BIBD are $t, b, r, k, \lambda$. For an experiment to be a BIBD, it must satisfy the following two conditions;

1.  $bk=rt$

2.  $\lambda(t-1)=r(k-1)$, this implies that; $\lambda=\frac{r(k-1)}{t-1}$. Note that $\lambda$ is the number of times each pair of treatment appear or occur together in the same block.

The model can be written as;

$$Y_{ij}=\mu+\tau_i+\beta_j+\epsilon_{ij}$$

where; i = 1,2,...,t, j = 1,2,...,r

$\mu$ - overall mean

$\tau_i$ - Effect of the ith treatment

$\beta_j$ - Effect of the jth block

$\epsilon_{ij}$ - Random error term

The sums of squares are partitioned as follows;

$$SS_{Total}=SS_{Block}+SS_{treat(adj)}+SS_{Error}$$

$$SS_{Total = \sum_i\sum_jY_{ij}^2}- \frac{Y_{..}^2}{N=(r\times t\space or\space b\times k)}$$

$$SS_{Block}=\frac{\sum_{j=1}^kY_{ij}^2}{k}-\frac{Y_{..}^2}{N}$$

$$SS_{treat(adj)}=k\frac{\sum_{i=1}^rQ_I}{\lambda t}$$

where $Q_i$ is the adjusted total for the ith treatment computed as;

$$Q_i=Y_{i.}-\frac{1}{k}\sum_jn_{ijY_{.j}}$$

i = 1,2,...,t

Then; $n_{ij}=1$ if treatment $i$ appears in block $j$ and $n_{ij}=0$ otherwise.

The ANOVA table is as follows;

+------------------------+----------------------------------+----------------------------------------------------+----------------------------------------------------------------------+
| Source of Variation    | df                               | SS                                                 | MS                                                                   |
+:======================:+:================================:+:==================================================:+:====================================================================:+
| $$                     | $$                               | $$                                                 | $$                                                                   |
|          Block         |                           b-1    |                                     SS_{Block}     |                                                       MS_{Block}     |
|           $$           |                           $$     |                                         $$         |                                                           $$         |
+------------------------+----------------------------------+----------------------------------------------------+----------------------------------------------------------------------+
| $$                     | $$                               | $$                                                 | $$                                                                   |
|       Treat(adj)       |                           t-1    |                                   SS_{treat(adj)}  |                                                     MS_{treat(adj)}  |
|           $$           |                           $$     |                                         $$         |                                                           $$         |
+------------------------+----------------------------------+----------------------------------------------------+----------------------------------------------------------------------+
| $$                     | $$                               | $$                                                 | $$                                                                   |
|          Error         |                         N-t-b+1  |                                     SS_{Error}     |                                                       MS_{Error}     |
|           $$           |                           $$     |                                         $$         |                                                           $$         |
+------------------------+----------------------------------+----------------------------------------------------+----------------------------------------------------------------------+
| $$                     | $$                               | $$                                                 |                                                                      |
|          Total         |                           N-1    |                                     SS_{Total}     |                                                                      |
|           $$           |                           $$     |                                         $$         |                                                                      |
+------------------------+----------------------------------+----------------------------------------------------+----------------------------------------------------------------------+

To test $H_o:\tau_i=0$ vs $H_1:\tau_i\ne0$, the test statistic is;

$$F_{calc}=\frac{MS_{treat(adj)}}{MS_{Error}}$$

We reject $H_0$ if $F_{calc}>F_{t-1,N-t-b+1}$ \newpage 

# 4. Partially Balanced Incomplete Design.

BIBD do not exist for all combination of parameters you might with to employ. PBIBD are designs whereby the number of times a given treatment occur will vary from treatment to treatment where some pairs of treatment appear together $\lambda_1$ times and others appear $\lambda_2$ times. Note $\lambda_1>\lambda_2$. This is the simplest of the PBIBD which had two associate classes; $\lambda_1$ and $\lambda_2$. For a design to be PBIBD, it must satisfy the following two conditions;

$$\sum_{i=1}^2n_i=t-1,\space n_1+n_2=t-1$$

$$\sum_{i=1}^2\lambda_in_i=r(k-1), \space \lambda_1n_1+\lambda_2n_2=r(k-1)$$

How to determine $P_{jk}^1$, you pick any two pair of treatments that are of first associate. Then to determine $P_{jk}^2$, you pick any two pairs of treatments which are second associate.

## Lattice Designs

Consider a balanced incomplete design with $t=k^2$ treatments arranged in $b=k(k+1)$ blocks with $k$ runs per blocks and $r=k+1$ replicates. Such a design is called a **balanced lattice**.

1.  A design for $k^2$ treatments in $2k$ blocks of $k$ runs with 2 replicates is called a **simple lattice**.

2.  A lattice design with $t=k^2$ treatments in $3k$ blocks grouped into 3 replicates, $r=3$ is called a **triple lattice design**.

3.  A lattice design for $t = k^2$ treatments in $4k$ blocks arranged in 4 replicates, $r=4$ is called a **quadruple lattice**.

4.  A lattice design for $t=k^3$ treatments in $k^2$ blocks of k runs is called a **cubic lattice design**.

5.  A lattice design for $t=k(k+1)$ treatments in $k+1$ blocks of size k is called a **rectangular lattice design**.
